---
title: 'EDS 223: assignment 4'
author: "Sofia Ingersoll"
date: "2023-12-2"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
```

## Overview

Marine aquaculture has the potential to play an important role in the global food supply as a more sustainable protein option than land-based meat production.[^1] [Gentry et al.](https://www.nature.com/articles/s41559-017-0257-9) mapped the potential for marine aquaculture globally based on multiple constraints, including ship traffic, dissolved oxygen, bottom depth .[^2]

[^1]: Hall, S. J., Delaporte, A., Phillips, M. J., Beveridge, M. & O'Keefe, M. Blue Frontiers: Managing the Environmental Costs of Aquaculture (The WorldFish Center, Penang, Malaysia, 2011).

[^2]: Gentry, R. R., Froehlich, H. E., Grimm, D., Kareiva, P., Parke, M., Rust, M., Gaines, S. D., & Halpern, B. S. Mapping the global potential for marine aquaculture. *Nature Ecology & Evolution*, 1, 1317-1324 (2017).

For this assignment, you are tasked with determining which Exclusive Economic Zones (EEZ) on the West Coast of the US are best suited to developing marine aquaculture for several species of oysters.\

Based on previous research, we know that oysters needs the following conditions for optimal growth:\

-   sea surface temperature: 11-30°C\
-   depth: 0-70 meters below sea level

##### Learning objectives:

-   combining vector/raster data\
-   resampling raster data\
-   masking raster data\
-   map algebra\

### Data

#### Sea Surface Temperature

We will use average annual sea surface temperature (SST) from the years 2008 to 2012 to characterize the average sea surface temperature within the region. The data we are working with was originally generated from [NOAA's 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php).

#### Bathymetry

To characterize the depth of the ocean we will use the [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area).[^3]

[^3]: GEBCO Compilation Group (2022) GEBCO_2022 Grid (<doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c>).

#### Exclusive Economic Zones

We will be designating maritime boundaries using Exclusive Economic Zones off of the west coast of US from [Marineregions.org](https://www.marineregions.org/eez.php).

## Assignment

Below is an outline of the steps you should consider taking to achieve the assignment tasks.

#### Prepare data (5 points)

To start, we need to load all necessary data and make sure it has the coordinate reference system.

-   load necessary packages and set path 
    -   I recommend using the [`here` package](https://here.r-lib.org/)

```{r message = FALSE}
# Loading libraries
library(sf)
library(tidyverse)
library(terra)
library(here)
library(tmap)
library(stars)
library(raster)
rm(list = ls())
setwd(here())
```

-   read in the shapefile for the West Coast EEZ (`wc_regions_clean.shp`)

```{r}
# Reading in shapefile
wc_regions_clean <- st_read("data/wc_regions/wc_regions_clean.shp",
                            quiet = TRUE)
# Checking CRS
crs(wc_regions_clean)

# Visualizing Data
plot(wc_regions_clean["rgn_id"])
```

-   read in SST rasters
    -   `average_annual_sst_2008.tif`\
    -   `average_annual_sst_2009.tif`\
    -   `average_annual_sst_2010.tif`\
    -   `average_annual_sst_2011.tif`\
    -   `average_annual_sst_2012.tif`

```{r}
# Reading in rasters
sst_08 <- read_stars("data/average_annual_sst_2008.tif") 
sst_09 <- read_stars("data/average_annual_sst_2009.tif")
sst_10 <- read_stars("data/average_annual_sst_2010.tif")
sst_11 <- read_stars("data/average_annual_sst_2011.tif")
sst_12 <- read_stars("data/average_annual_sst_2012.tif")

# Checkng CRS, shows evidence ofno projection
crs(sst_08)
```

-   combine SST rasters into a raster stack

```{r}
# Combining SST rasters into raster stack

# object tifs names are stored in 
files_list <- c("data/sst/average_annual_sst_2008.tif", 
                "data/sst/average_annual_sst_2009.tif", 
                "data/sst/average_annual_sst_2010.tif",
                "data/sst/average_annual_sst_2011.tif", 
                "data/sst/average_annual_sst_2012.tif")
# SpatRaster
sst_stack <- rast(files_list) %>% 
  project("EPSG:4326")

# Confirm all tifs in stack and projected correctly
sst_stacked                        

plot(sst_stack)
#---------------------------------------------------------------------
# large rasterlayers & large list 
# Creates a storing list of tifs
file_list <- list.files(
  here("data/sst"),
  pattern = ".tif",
  full.names = TRUE
)

# reference raster for resampling
ref_sst <- raster(file_list[1])

# raster stack as SpatRaster object & projecting it
sst_stacked <- list()

for (file in file_list) {
  sst <- raster(file)
  resampled_sst <- resample(sst, ref_sst) 
  sst_crop <- crop(resampled_sst, extent(ref_sst))
  sst_stacked[[file]] <- sst_crop
}

# Visualizing the components of the ratster stack
plot(sst_stacked)
```

-   read in bathymetry raster (`depth.tif`)

```{r}
# Reading in depth tif spatraster
depth <- read_stars("data/depth.tif")

# Checking crs, provides evidence that this tif is not projected
crs(depth)

# Projecting CRS
depth <- rast(depth) %>% 
 project("EPSG:4326")
#-------------------------------------------------
# Projecting CRS
depth <- raster(depth) %>% 
 project("EPSG:4326")

# Confirming projection succuessful
crs(depth)

# Visualizing depth
plot(depth)
```

-   check that data are in the same coordinate reference system\
    -   reproject any data not in the same projection

```{r include=TRUE, warning=FALSE, messages=FALSE}
# Checking CRS
glimpse(crs(wc_regions_clean))
glimpse(crs(sst_stack))
glimpse(crs(depth))
```

#### Process data (10 points)

Next, we need process the SST and depth data so that they can be combined. In this case the SST and depth data have slightly different resolutions, extents, and positions. We don't want to change the underlying depth data, so we will need to resample to match the SST data using the nearest neighbor approach.

-   find the mean SST from 2008-2012

```{r}
# Finding the mean

# Mean of the overall stack 
mean_sst_stack <- mean(sst_stack, na.rm = TRUE)

mean_sst_stack
#-------------------------------------------------------------------------- 
# ideas but unsure if needed

# Mean for each year in layers
# not the right data type rn
#mean_sst_summary <- summarize(mean_sst_08 = mean(sst_08$y),
 #                             mean_sst_09 = mean(sst_09$y),
  #                            mean_sst_10 = mean(sst_10$y), 
   #                           mean_sst_11 = mean(sst_11$y),
    #                          mean_sst_12 = mean(sst_12$y))
# averaging the mean outcomes
#sst_means <- mean_sst_summary %>% 
 # mean(mean_sst_08, mean_sst_09, mean_sst_10, mean_sst_11, mean_sst_12)

```

-   convert SST data from Kelvin to Celsius\
    -   hint: subtract by 273.15

```{r}
# Converting from Kelvin to Celcius

# need to find function that works on SpatRaster
# mutate and transmute not working

#mean_sst_stack <- mean_sst_stack %>% 
 # transmute(mean_C = mean_sst_stack$mean - 273.15)

mean_sst_stack_c <- mean_sst_stack - 273.15 
mean_sst_stack_c
```

-   crop depth raster to match the extent of the SST raster
    -   note: the resolutions of the SST and depth data do not match

```{r}
depth
```

-   resample the NPP data to match the resolution of the SST data using the nearest neighbor approach

```{r}
# resampling depth to match mean_sst_stack resolution using nearest nbr
depth_resampled <- resample(depth, mean_sst_stack_c, method = "near")

# boolean test to see if resolutions match
resolution(depth_resampled) == resolution(mean_sst_stack_c)
```

-   check that the depth and SST match in resolution, extent, and coordinate reference system\
    -   hint: can the rasters be stacked?

```{r include=TRUE}
# confirming last step successful with stack
depth_sst <- rast(c(mean_sst_stack_c, depth_resampled))

plot(depth_sst)
```

#### Find suitable locations (20)

In order to find suitable locations for marine aquaculture, we'll need to find locations that are suitable in terms of both SST and depth.

-   reclassify SST and depth data into locations that are suitable for Lump sucker fish\
    -   hint: set suitable values to `1` and unsuitable values to `NA`\
-   find locations that satisfy both SST and depth conditions\
    -   hint: create an overlay using the `lapp()` function multiplying cell values\

```{r include=TRUE}

```

#### Determine the most suitable EEZ (20 points)

We want to determine the total suitable area within each EEZ in order to rank zones by priority. To do so, we need to find the total area of suitable locations within each EEZ.

-   select suitable cells within West Coast EEZs\
-   find area of grid cells\
-   find the total suitable area within each EEZ\
    -   hint: it might be helpful to rasterize the EEZ data\
-   find the percentage of each zone that is suitable\
    -   hint it might be helpful to join the suitable area by region onto the EEZ vector data\

```{r include=TRUE}

```

#### Visualize results (5 points)

Now that we have results, we need to present them!

Create the following maps:

-   total suitable area by region\
-   percent suitable area by region\

Include:

-   legible legends\
-   updated color aesthetics\
-   basemap\

```{r include=TRUE}

```

#### Broaden your workflow! (40 points)

Now that you've worked through the solution for one group of species, let's update your workflow to work for other species. Please create a function that would allow you to reproduce your results for other species. Your function should be able to do the following:\

-   accept temperature and depth ranges and species name as inputs\
-   create maps of total suitable area and percent suitable area per EEZ with the species name in the title\

Run your function for a species of your choice! You can find information on species depth and temperature requirements on [SeaLifeBase](https://www.sealifebase.ca/search.php). Remember, we are thinking about the potential for marine aquaculture, so these species should have some reasonable potential for commercial consumption.
